Shallow EDSLs and Object-Oriented Programming

### 2017-01-17 paper/sections/Introduction.tex
JFP format

\begin{quote}
\emph{With a deep embedding, terms in the DSL are implemented simply to
construct an abstract syntax tree (AST), which is subsequently
transformed for optimization and traversed for evaluation. With a
shallow embedding, terms in the DSL are implemented directly by
their semantics, bypassing the intermediate AST and its traversal.}\cite{gibbons2014folding}
\end{quote}

@@

The main goal of our paper is to show the close relationship between 
shallow embeddings and OOP, and argue that OOP languages have
advantages for the implementation of shallow embeddings. 
It is perhaps partly due to those advantages that some authors 
have considered  {\sc Composite}-based implementations to be deep
embeddings. However, we argue that the OOP
mechanisms do not change the essence of the implementation, which is
still shallow (i.e. using procedural abstraction). 
To understand how
OOP is helpful for shallow embeddings, we should first look at the
limitations of shallow embeddings commonly found in the literature:
\begin{enumerate}

\item {\bf Single Interpretation} An often stated limitation of
  shallow EDSLs is that they only support one
  interpretation.

\item {\bf No Transformations} Another commonly stated limitation 
of shallow embeddings is the lack of support for transformations,
preventing optimizations.

\end{enumerate}

We show that OOP abstractions, including \emph{
  inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address those problems. For the first problem, we can
employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanisms enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
For the second problem, we show that transformations can be encoded 
with recursive objects. Interestingly enough, for transformations it
is possible to port back the OO solution to
Haskell.

>>>>

\begin{quote}
With a deep embedding, terms in the DSL are implemented simply to
construct an abstract syntax tree (AST), which is subsequently
transformed for optimization and traversed for evaluation. With a
shallow embedding, terms in the DSL are implemented directly by
their semantics, bypassing the intermediate AST and its traversal.\cite{gibbons2014folding}
\end{quote}

@@

The main goal of our paper is to show the close relationship between 
shallow embeddings and OOP, and argue that OOP languages have
advantages for the implementation of shallow embeddings. 
It is perhaps partly due to those advantages that some authors 
have considered  {\sc Composite}-based implementations to be deep
embeddings. However, we argue that the OOP
mechanisms do not change the essence of the implementation, which is
still shallow (i.e. using procedural abstraction).
An often stated limitation of shallow EDSLs is that they only support \emph{single} interpretation.
We show that OOP abstractions, including \emph{inheritance}, \emph{subtyping} and \emph{type-refinement}, are
helpful to address this problem.
We can employ a recently proposed design pattern~\cite{eptrivially16}, which provides a simple
solution to the \emph{Expression Problem}~\cite{expPb} in OOP languages. Thus
using just standard OOP mechanisms enables \emph{multiple modular
  interpretations} to co-exist and be combined in shallow embeddings.
